---
title: 'Health Disparities: U.S.-Born vs Foreign-Born Populations'
subtitle: "NHIS 2023 Analysis Portfolio Project"
author: "Nadia Shureih"
date: "April 3rd, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
# Environment setup
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = 'center'
)
```

# Introduction

As a recent public health graduate with data analytics training, I conducted this analysis to explore health disparities using the **2023 National Health Interview Survey (NHIS)**. This project demonstrates my ability to:

-   Clean and analyze complex survey data

-   Apply appropriate weighting techniques

-   Communicate findings effectively to public health audiences

## Study Population

Comparing three groups:

1.  **U.S.-born residents** (baseline group)

2.  **Foreign-born naturalized citizens**

3.  **Foreign-born non-citizens**

*Note: "Foreign-born" includes all individuals born outside the U.S., regardless of immigration status. NHIS does not collect undocumented status.*

## Public Health Relevance

Understanding these disparities is critical for:

-   Identifying at-risk populations

-   Informing culturally-appropriate interventions

-   Evaluating healthcare access policies

-   Addressing social determinants of health

# Data Preparation

## Methodology Overview

The NHIS uses a complex sampling design, so all analyses incorporate survey weights (wtfa_a) to ensure nationally representative estimates. I followed these steps:

1.  Data cleaning and recoding

2.  Variable transformation

3.  Weighted analysis

4.  Visualization

```{r load-data}
# Load necessary packages
library(tidyverse)   # For data cleaning and visualization
library(janitor)     # For cleaning variable names
library(survey)      # For weighted analysis
library(gtsummary)   # For professional tables
library(ggthemes)    # For professional ggplot themes
library(scales)      # For better axis formatting

# Load data with error handling
tryCatch({
  nhis_data <- read_csv("Project/adult23.csv") %>% 
    clean_names() %>%  
    select(
      id = hhx,
      weight = wtfa_a,
      sex = sex_a,
      age = agep_a,
      citizenship = citznstp_a,
      nativity = natusborn_a,
      years_in_us = yrsinus_a,
      health_insurance = cover_a,
      depression = depev_a,
      anxiety = anxev_a
    )
}, error = function(e) {
  stop("Data loading failed. Please check file path and permissions.")
})
```

## Cleaning the Data

Here's how I prepared the variables for analysis:

```{nhis_clean <- nhis_data %>%}
  mutate(
    # Create population groups
    population_group = case_when(
      nativity == 1 ~ "U.S.-born",
      citizenship == 2 ~ "Naturalized citizen",
      citizenship == 3 ~ "Non-citizen",
      TRUE ~ "Other"
    ) %>% factor(levels = c("U.S.-born", "Naturalized citizen", "Non-citizen")),
    
    # Create binary insurance variable (1 = uninsured, 0 = insured)
    uninsured = ifelse(health_insurance == 4, 1, 0),
    
    # Mental health indicator
    mental_health_issue = ifelse(depression == 1 | anxiety == 1, 1, 0),
    
    # Time in U.S. categories
    years_in_us_group = case_when(
      nativity == 1 ~ "U.S.-born",
      years_in_us < 5 ~ "0-4 years",
      years_in_us < 10 ~ "5-9 years",
      years_in_us >= 10 ~ "10+ years",
      TRUE ~ "Missing"
    )
  ) %>%
  filter(population_group != "Other",
         !is.na(weight))
```

## Survey Design ObjectKey Findings

```{r}
# Create survey design object for proper weighting
nhis_design <- svydesign(
  ids = ~1,
  weights = ~weight,
  data = nhis_clean
)
```

# Descriptive Statistics

```{r}
# Create demographic table
nhis_clean %>%
  select(population_group, sex, age_group, insurance_status, mental_health_issue) %>%
  tbl_summary(
    by = population_group,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_categorical() ~ 1),
    missing = "no"
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table 1.** Demographic Characteristics by Nativity Status")
```

# Key Findings

## 1. Health Insurance Coverage Disparities

```{# Calculate weighted proportions with confidence intervals}
insurance_svy <- svyby(~insurance_status == "Uninsured", ~population_group, 
                      nhis_design, svymean, na.rm = TRUE, vartype = "ci")

# Rename columns for clarity
insurance_plot_data <- insurance_svy %>%
  rename(percent = `insurance_status == "Uninsured"`,
         lci = ci_l,
         uci = ci_u) %>%
  mutate(percent = percent * 100,
         lci = lci * 100,
         uci = uci * 100)

# Create visualization
ggplot(insurance_plot_data, 
       aes(x = population_group, y = percent, fill = population_group)) +
  geom_col(alpha = 0.8) +
  geom_errorbar(aes(ymin = lci, ymax = uci), width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", percent)), 
            vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Figure 1. Uninsured Rates by Nativity Status",
    subtitle = "Foreign-born non-citizens have significantly higher uninsured rates\nError bars represent 95% confidence intervals",
    x = "",
    y = "Percentage Uninsured",
    caption = "Source: 2023 NHIS, weighted estimates"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"))
```

**What this shows:** - Naturalized citizens have similar insurance rates to U.S.-born residents (about 90% covered) - Non-citizens are much more likely to be uninsured (only 63.3% insured).

## 2. Mental Health by Time in U.S.

```{r mental-health-plot}
# Filter to foreign-born only
foreign_born <- nhis_clean %>%
  filter(nativity == 2 & years_in_us_group != "Missing")

# Calculate rates
mental_health_rates <- foreign_born %>%
  group_by(years_in_us_group) %>%
  summarize(
    rate = weighted.mean(mental_health_issue, weight),
    .groups = "drop"
  )

# Create plot
ggplot(mental_health_rates, aes(x = years_in_us_group, y = rate)) +
  geom_col(fill = "#4E79A7") +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -0.5) +
  labs(
    title = "Mental Health Issues Increase with Time in U.S.",
    subtitle = "Among foreign-born residents",
    x = "Years in United States",
    y = "Percentage with Depression/Anxiety"
  ) +
  theme_minimal()
```

**Key takeaway:** Mental health issues appear more common among those who've lived in the U.S. longer.

Possible explanations:

\- Stress of acculturation

\- Loss of original support networks

\- Changing environmental factors

# Limitations

While this analysis provides useful insights, there are important limitations:

1.  **The NHIS doesn't distinguish between:**
    -   Legal permanent residents
    -   Temporary visa holders
    -   Undocumented individuals
2.  **Self-reported data** may underreport certain conditions due to:
    -   Stigma (especially mental health)
    -   Lack of diagnosis
3.  **Sample size limitations** for smaller subgroups

# Conclusion

This analysis reveals clear health disparities between population groups:

1.  **Insurance access** remains a major challenge for non-citizens
2.  **Mental health support** may need to evolve for long-term foreign-born residents
3.  **Future research** should examine specific subpopulations

As I continue developing my data skills, I'd like to explore:

\- Adding more social determinants (housing, employment)

\- Incorporating geographic differences

```{r session-info, include=FALSE}
sessionInfo()
```
